---
title: "The Community Eligibility Provision: Food Insecurity in Durham Public Schools"
author: "Julia Donheiser"
date: "11/2/2018"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE)

library(readxl) # read spreadsheets
library(dplyr) # wrangling
library(ggplot2) # graphing
library(tidyr) # reshaping
library(stringr) # string manipulation
library(data.table) # working with tables
library(cowplot) # layout plots
library(knitr) # nice tables
library(directlabels) # label plots

df = read.csv("./data/master.csv")
df$cep = df$cep %>% as.factor()
df$cep_long = df$cep_long %>% as.factor()
```

## Exploratory Data Analysis
```{r}
# debt over time, excluding non-standard and inactive schools
df %>%
  group_by(year) %>%
  summarise(total_debt = sum(unpaid, na.rm = TRUE),
            mean_debt_per_student = mean(debt_per_student, na.rm = TRUE),
            fullprice_lunches = total_debt/2.90)

ggplot(df[df$year != '2010-11',], aes(year, unpaid)) + 
  geom_bar(stat="sum") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))
  

# percent change since 2014-15, first CEP year
mean(df$debt_per_student[df$ts==2017])/mean(df$debt_per_student[df$ts==2014])
```
At the end of the 2017-18 academic year, DPS had over \$211,000 in school lunch debt. That's over 72,000 unpaid lunches, with an average of \$5.10 of debt per student. It's also the most debt the school district has seen in the past eight years.

```{r, fig.width=5, fig.height=3, fig.align="center"}
# histogram
hist(df$debt_per_student, main = "", xlab = "debt-per-student")
```
Most schools have less than \$5 of lunch debt per student. In Durham, a full-priced lunch costs \$2.90, and a reduced-price lunch cost \$0.40, according to the Durham Public Schools \href{https://www.dpsnc.net/site/handlers/filedownload.ashx?moduleinstanceid=196&dataid=173&FileName=2018-19%20FR%20Lunch%20App_Eng.pdf}{website}. That's about two unpaid full-priced lunches per student, or just over 12 unpaid reduced-price lunches per student. For the rest of my EDA, I'll delve into which schools have more debt and whether we can find systematic issues. I'll also be looking at schools that are part of the Community Eligibility Provision, which means all students receive free lunch.

### Debt and Demographics
```{r, fig.width=12, fig.height=4, fig.align="center"}
# scatterplot of debt vs. pct needy
ggplot(df[df$cep == 0,], aes(pct_free_reduced, debt_per_student)) +
  geom_point(aes(col = cep_long)) +
  xlab("percent students with free and reduced lunch") +
  ylab("debt-per-student") +
  geom_smooth(method = "lm", se = FALSE)

# scatterplot of debt vs. pct black
g1 = ggplot(df[df$cep == 0,], aes(pct_black, debt_per_student)) +
  geom_point(aes(col = cep_long)) +
  xlab("Percent Black") +
  ylab("debt-per-student") +
  geom_smooth(method = "lm", se = FALSE)

# scatterplot of debt vs. pct hispanic
g2 = ggplot(df[df$cep == 0,], aes(pct_hispanic, debt_per_student)) +
  geom_point(aes(col = cep_long)) +
  xlab("Percent Hispanic") +
  ylab("debt-per-student") +
  geom_smooth(method = "lm", se = FALSE)

# scatterplot of debt vs. pct white
g3 = ggplot(df[df$cep == 0,], aes(pct_white, debt_per_student)) +
  geom_point(aes(col = cep_long)) +
  xlab("Percent White") +
  ylab("debt-per-student") +
  geom_smooth(method = "lm", se = FALSE)

plot_grid(g1, g2, g3, ncol = 3)

ggplot(df[df$cep == 0,], aes(pct_minority, debt_per_student)) +
  geom_point(aes(col = cep_long)) +
  xlab("Percent Minority") +
  ylab("debt-per-student") +
  geom_smooth(method = "lm", se = FALSE)

g1 = ggplot(df, aes(pct_white, pct_black)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent White") +
  ylab("Percent Black")

g2 = ggplot(df, aes(pct_white, pct_hispanic)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent White") +
  ylab("Percent Hispanic")

g3 = ggplot(df, aes(pct_black, pct_hispanic)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Black") +
  ylab("Percent Hispanic")

plot_grid(g1, g2, g3, ncol = 3)

ggplot(df, aes(pct_minority, pct_white)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Minority") +
  ylab("Percent White")
```

There is a weak positive correlation between the percentage of students who have free and reduced price lunch and the debt-per-student at each school. It looks like schools with CEP status generally have a higher percentage of students on free/reduced lunch. As for demographics, there is also a weak positive correlation between the percentage of black students and debt-per-student. Conversely, there is a weak negative correlation between the percent of white students and debt-per-student. There doesn't appear to be a much of a relationship between the percentage of hispanic students and debt-per-student. That being said, if we look at the total percentage of minority students --- the percentage of black \textit{and} hispanic students --- there is a positive correlation with debt-per-student. Schools that CEP status are also starkly segregated, with very few white students and primarily black and hispanic students.

### Debt over time
```{r debt_over_time, fig.width=8, fig.height=4}
# debt by school over time
p1 = ggplot(df, aes(year, unpaid, group = school_no, col = cep_long)) +
  geom_point(aes(col=cep)) +
  geom_line() +
  ylab("total debt") +
  ggtitle("Total Debt by School") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12)) +
  facet_wrap(~ cep_long, nrow = 2)

# debt/student over time
p2 = ggplot(df, aes(year, debt_per_student, group = school_no, col = cep_long)) +
  geom_point(aes(col=cep)) +
  geom_line() +
  ylab("debt per student") +
  ggtitle("Debt per Student by School") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12)) +
  facet_wrap(~ cep_long, nrow = 2)

plot_grid(p1, p2)
```

Schools with CEP status generally had a large amount of debt until the 2014-15 school year, when they gained CEP status. That being said, other schools with similar amounts of debt did not gain CEP status.

### Demographics and need
```{r, fig.width=12, fig.height=4, fig.align="center"}
g1 = ggplot(df[df$cep == 0,], aes(pct_white, pct_free_reduced)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent White") +
  ylab("percent students with free and reduced lunch")

g2 = ggplot(df[df$cep == 0,], aes(pct_hispanic, pct_free_reduced)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Hispanic") +
  ylab("percent students with free and reduced lunch")

g3 = ggplot(df[df$cep == 0,], aes(pct_black, pct_free_reduced)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Black") +
  ylab("percent students with free and reduced lunch")

g4 = ggplot(df[df$cep == 0,], aes(pct_minority, pct_free_reduced)) +
  geom_point(aes(col = cep_long)) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Percent Minority") +
  ylab("percent students with free and reduced lunch")

plot_grid(g1, g2, g3, g4, ncol = 4)
```

We can also see that race is a proxy for need. The percentage of black and hispanic schools have a strong, positive correlation with the percentage of students on free and reduced price lunch. For white students, this relationship is strong and negative.

### Locale and debt, race
```{r, fig.width=12, fig.height=4, fig.align="center"}
b1 = ggplot(df[df$cep == 0,], aes(urban_centric_locale, unpaid)) +
  geom_boxplot()

b2 = ggplot(df[df$cep == 0,], aes(urban_centric_locale, debt_per_student)) +
  geom_boxplot()

plot_grid(b1, b2, ncol = 2)
```
There don't seem to be major differences in debt by location.

```{r, fig.width=12, fig.height=4, fig.align="center"}
b1 = ggplot(df, aes(urban_centric_locale, pct_white)) +
  geom_boxplot() +
  ylim(c(0,1))

b2 = ggplot(df, aes(urban_centric_locale, pct_black)) +
  geom_boxplot() +
  ylim(c(0,1))

b3 = ggplot(df, aes(urban_centric_locale, pct_hispanic)) +
  geom_boxplot() +
  ylim(c(0,1))

b4 = ggplot(df, aes(urban_centric_locale, pct_minority)) +
  geom_boxplot() +
  ylim(c(0,1))

plot_grid(b1, b2, b3, b4, ncol = 4)
```
On average, we also don't see any major differences in race by locale. However, there is a subset of rural schools that seem to primarily be white.

### Debt and grade
```{r}
df %>%
  filter(cep == 0) %>%
  group_by(grade_clean) %>%
  summarise(mean_debt = mean(unpaid),
            mean_dps = mean(debt_per_student))

debtByGrade = df %>%
  filter(cep == 0) %>%
  group_by(year, grade_clean) %>%
  summarise(mean_dps = mean(debt_per_student))

ggplot(debtByGrade, aes(year, mean_dps, group=grade_clean, col=grade_clean)) +
  geom_point() +
  geom_line() +
  ylab("Debt per Student") +
  ggtitle("Debt per Student by Grade Level") +
  scale_y_continuous(labels = scales::dollar) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))
```
On average, high schools have the most total debt. However, middle schools, on average, have the most debt per student.

```{r, fig.width=12, fig.height=4, fig.align="center"}
b1 = ggplot(df[df$cep == 0,], aes(grade_clean, unpaid)) +
  geom_boxplot()

b2 = ggplot(df[df$cep == 0,], aes(grade_clean, debt_per_student)) +
  geom_boxplot()

plot_grid(b1, b2, ncol = 2)
```

## CEP Schools
In this section, I explore the differences between schools that gained CEP status and schools that did not.

### Eligibility
```{r}
# CEP as factor
df$cep = df$cep %>% as.character() %>% as.numeric()

# Eligibility and participation counts
df %>%
  summarise(eligible = sum(elig, na.rm = TRUE),
            participating = sum(participating, na.rm = TRUE),
            near_eligible = sum(near_elig, na.rm = TRUE))
```

### Eligible but non-participating CEP schools
```{r}
# total debt for eligible but not participating schools
df %>%
  filter(ts==2017) %>%
  group_by(elig, participating) %>%
  summarise(n = n(),
            unpaid = sum(unpaid),
            mean_dps = round(mean(debt_per_student),2),
            avg_free_reduced = round(mean(pct_free_reduced),2),
            avg_pct_minority = round(mean(pct_minority),2))

# percentage of total 2017-18 debt from CEP eligible but not participating schools
sum(df$unpaid[(df$ts==2017) & (df$elig==1) & (df$participating==0)])/sum(df$unpaid[df$ts==2017])
```
Last year, debt was concentrated in schools that were not eligible to participate in the CEP program. However, eligible but non-participating CEP schools accounted for nearly 40 percent of all debt in Durham county.

### CEP Status and Grade Level
```{r}
df %>%
  filter(ts==2017) %>%
  group_by(grade_clean, elig, participating) %>%
  summarise(n = n(),
            unpaid = sum(unpaid),
            mean_dps = round(mean(debt_per_student),2)) %>%
  arrange(-mean_dps)

df %>%
  filter(ts==2017) %>%
  group_by(grade_clean, elig, participating) %>%
  summarise(n = n(),
            unpaid = sum(unpaid),
            mean_dps = round(mean(debt_per_student),2)) %>%
  arrange(-unpaid)
```
Last year, all but one of the CEP schools were elementary schools. We also see that no middle schools participated in the CEP program, but the six that were eligible & non-participating had the most debt per student--nearly twice that of eligible but non-participating elementary schools. Eligible and non-participating high-schools, however, had relatively low debt per student at just $2.90 of unpaid meals per student. The overall debt for those schools was also fairly low.

```{r}
# indicator for whether school was eligible to participate in 2017-18
el17 = df$school_no[(df$elig==1) & (df$ts==2017)]
df$elig_long = 0
df$elig_long[df$school_no %in% el17] = 1

ggplot(df[df$elig_long==1,], aes(year,debt_per_student, group=school_name, col=cep_long)) +
  geom_point() +
  geom_line() +
  ylab("debt per student") +
  xlab("year") +
  ggtitle("debt per student among schools that were cep eligible in 2017-18") +
  geom_dl(aes(label=school_name), method = list("last.points")) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_discrete(limits=c(levels(df$year),"")) + 
  annotate("rect", xmin = "2013-14", xmax = "2014-15", ymin = -Inf, ymax = Inf,
             alpha = 0.2, fill = "pink") +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))

ggplot(df[df$elig_long==1,], aes(year,unpaid, group=school_name, col=cep_long)) +
  geom_point() +
  geom_line() +
  ylab("total debt") +
  xlab("year") +
  ggtitle("school lunch debt among schools that were cep eligible in 2017-18") +
  geom_dl(aes(label=school_name), method = list("last.points")) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_discrete(limits=c(levels(df$year),"")) + 
  annotate("rect", xmin = "2013-14", xmax = "2014-15", ymin = -Inf, ymax = Inf,
             alpha = 0.2, fill = "pink") +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))

ggplot(df[df$cep_long==1,], aes(year,unpaid, group=school_name)) +
  geom_point(aes(col=cep)) +
  geom_line() +
  ylab("total debt") +
  xlab("year") +
  ggtitle("school lunch debt among schools that have participated in CEP") +
  geom_dl(aes(label=school_name), method = list("first.points")) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_discrete(limits=c("","",levels(df$year))) +
  annotate("rect", xmin = "2013-14", xmax = "2014-15", ymin = -Inf, ymax = Inf,
             alpha = 0.2, fill = "pink") +  
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))

ggplot(df[df$cep_long==1,], aes(year,debt_per_student, group=school_name)) +
  geom_point(aes(col=cep)) +
  geom_line() +
  ylab("debt per student") +
  xlab("year") +
  ggtitle("debt per student among schools that have participated in CEP") +
  geom_dl(aes(label=school_name), method = list("first.points")) +
  scale_y_continuous(labels = scales::dollar) +
  scale_x_discrete(limits=c("","",levels(df$year))) + 
  annotate("rect", xmin = "2013-14", xmax = "2014-15", ymin = -Inf, ymax = Inf,
             alpha = 0.2, fill = "pink") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))
```

### Percent Minority
```{r}
df %>%
  group_by(cep_long) %>%
  summarise(mean_minority = mean(pct_minority))
```
CEP schools also have a higher concentration of minority students, on average.

### Debt before CEP status
```{r}
# TODO: quantify debt among CEP schools the year before they enrolled in the program
# this is a slightly inaccurate representation of total debt
df %>%
  filter(ts < 2015) %>%
  group_by(cep_long) %>%
  summarise(mean_dps = mean(debt_per_student),
            mean_debt = mean(unpaid))
```
CEP schools also had, on average, higher debt and more debt per student than non-CEP schools prior to CEP assignments.

## Individual school analysis
### Neal Middle
```{r}
df %>%
  filter(school_name == "Neal") %>%
  select(year, unpaid, debt_per_student)
```

## CEP vs. ISP comparison
```{r}
df$isp_diff = df$pct_free_reduced - df$isp*1.6

df %>%
  filter(ts >= 2015,
         eligible == 1,
         participating == 0) %>%
  group_by(school_name,
           school_no) %>%
  summarise(diff = mean(isp_diff)*100) %>%
  arrange(-diff)
```

