---
title: "Unpaid Lunch Debt in Durham, NC"
author: "Julia Donheiser"
date: "11/2/2018"
output: pdf_document
urlcolor: blue
---

## Introduction
When families can’t afford to pay for student lunches, school districts foot the bill. But with major cuts to educational funding in North Carolina—where some schools don’t even have enough funds to pay for students’ textbooks—this means school districts can wrack up tens of thousands of dollars in debt. In Durham, students with five or more unpaid lunches only receive a juice and a sandwich instead of a hot lunch. This lends its way to “lunch shaming”, where students who can’t afford pay skip the meal altogether to avoid the embarrassment of eating a cold lunch. This is a major issue, since student performance in school is directly tied to access to quality food.

### Data Sources
* End-of-Year unpaid meal data from James Keaton, director of child nutrition services at DPS.
* All free/reduced price lunch data was obtained from \href{http://www.ncpublicschools.org/fbs/resources/data/}{ncpublicschools.org}
* 2010-11 through 2015-16 demographic data was obtained from the \href{https://nces.ed.gov/ccd/elsi/tableGenerator.aspx}{NCES ELSI table generator}, code 91803
* 2017-18 ADM data from ncpublicschools.org's \href{http://www.ncpublicschools.org/fbs/accounting/data/}{Average Daily Membership and Membership Last Day by School}
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE)

library(readxl) # read spreadsheets
library(dplyr) # wrangling
library(ggplot2) # graphing
library(tidyr) # reshaping
library(stringr) # string manipulation
library(data.table) # working with tables
library(cowplot) # layout plots

# get all older files, .xls format
old.dat = list.files("./data") %>%
  grep("freereduced.xls$", ., value = TRUE)

# read files as table
dat = lapply(paste("./data/",old.dat,sep=""), function(x) read_xls(x))

# clean empty rows
head(dat[[1]])
colnames(dat[[1]]) = c("lea_no","lea_name","school_no","school_name","adm","reduced","free","pct_needy","grade")
dat[[1]] = dat[[1]][2:dim(dat[[1]])[1],]
head(dat[[2]])
colnames(dat[[2]]) = c("lea_no","lea_name","school_no","school_name","adm","reduced","free","pct_needy","grade")
dat[[2]] = dat[[2]][5:dim(dat[[2]])[1],]
head(dat[[3]])
colnames(dat[[3]]) = c("lea_no","lea_name","school_no","school_name","adm","reduced","free","pct_needy","grade")
dat[[3]] = dat[[3]][7:dim(dat[[3]])[1],]
head(dat[[4]])
colnames(dat[[4]]) = c("lea_no","lea_name","school_no","school_name","adm","reduced","free","pct_needy","grade")
dat[[4]] = dat[[4]][7:dim(dat[[4]])[1],]
head(dat[[5]])
colnames(dat[[5]]) = c("lea_no","lea_name","school_no","school_name","cep","adm","free","reduced","pct_needy","pct_needy_mlt")

# get newer files, .xlsx format
new.dat = list.files("./data") %>%
  grep("freereduced.xlsx$", ., value = TRUE)
dat2 = lapply(paste("./data/",new.dat,sep=""), function(x) read_xlsx(x))

# check for cleaning
head(dat2[[1]])
colnames(dat2[[1]]) = c("lea_no","lea_name","school_no","school_name","cep","adm","free","reduced","pct_needy","pct_needy_mlt")
head(dat2[[2]])
colnames(dat2[[2]]) = c("lea_no","lea_name","school_no","school_name","cep","adm","free","reduced","pct_needy")

# join all data
dat[6:7] = dat2

# transform to data frames
dat = lapply(dat, function(x) data.frame(x))

# remove totals (last row)
dat = lapply(dat, function(x) x[-nrow(x),])

# add id for year
yrs = c(old.dat, new.dat) %>% substr(., 0, 7) # get year to bind as column
for(i in 1:length(dat)) {
  dat[[i]]$year = yrs[i]
}

# merge
df = do.call(plyr::rbind.fill, dat)

# indicator for CEP data
df$cep = ifelse(is.na(df$cep), 0, 1)

# filter to durham county
df = df[grepl("Durham", df$lea_name, ignore.case = TRUE),]

# load lunch debt data
debt = read.csv("./data/meal_debt.csv")
debt = debt[1:nrow(debt)-1,1:10] # remove totals at bottom, only full year data
colnames(debt)[1:2] = c("school_no", "school_name")
colnames(debt)[3:9] = yrs # reform years
colnames(debt)[10] = "2017-18"

# reshape wide to long
debt = gather(debt, year, unpaid, 3:10)

# convert to integer
debt$unpaid = debt$unpaid %>%
  gsub("\\$","",.) %>%
  gsub("\\,","",.) %>%
  as.numeric()

# convert to factor for joining
df$school_no = as.factor(df$school_no)
debt$school_no = as.factor(debt$school_no)

# join unpaid lunch with free/reduced lunch data
df = left_join(debt, df, by = c("year","school_no"))

# dedupe school name cols
colnames(df)[which(colnames(df)=="school_name.x")] = "school_name"
colnames(df)[which(colnames(df)=="school_name.y")] = "school_name2"

# year as factor
df$year = as.factor(df$year)

# character to numeric
df[c("adm","reduced","free","pct_needy")] = sapply(df[c("adm","reduced","free","pct_needy")], as.numeric)

# add 2017-18 adm data
adm17 = read.csv("./data/2017-18adm.csv")

# filter to durham
adm17 = adm17[grepl("Durham", adm17$LEA.Name, ignore.case = TRUE),]

# only need adm and school number
adm17 = adm17[c("SCH","Total")]

# total as numeric
adm17$Total = as.numeric(adm17$Total)

# fill 2017-18 adm data
df$adm[df$year=="2017-18"] = adm17$Total[df$school_no[df$year=="2017-18"]]

# add demographic data
dem = read.csv("./data/nces_demographics.csv")

# remove state, only NC
dem = dem[-c(2)]

# reshape to get year as variable
dem = gather(dem, var, val, 2:73)

# remove id rows
spchars = c("†","‡","–")
dem = dem[!grepl(paste(spchars,collapse="|"),dem$School.Name),]

# add year as variable
dem$year = dem$var %>%
  substr(., nchar(.)-6, nchar(.)) %>% 
  gsub("\\.","\\-",.)

# remove year and "public school" from var name
dem$var = dem$var %>%
  substr(., 0, nchar(.)-24)

# set non-numeric values
dem$val[dem$val %in% spchars] = NA

# set alternative zeros to 0
dem$val[dem$val %in% c("=0","=000")] = 0

# long to wide
dem = dcast(dem, School.Name + year ~ var, value.var = "val")

## WHY ISN'T THIS WORKING!

# edit column names
colnames(dem) = colnames(dem) %>%
  gsub("\\.+","_",.) %>%
  tolower(.)

colnames(dem) = colnames(dem) %>% gsub("\\_students","",.)

# filter to durham
dem = dem[grepl("Durham", dem$county_name, ignore.case = TRUE),]
```

## Exploratory Data Analysis
```{r debt_over_time, fig.width=8, fig.height=4}
# normalize debt by adm
df$debt_per_student = df$unpaid/df$adm

# debt by school over time
p1 = ggplot(df, aes(year, unpaid, group = school_no, col = school_no)) +
  geom_point() +
  geom_line() +
  ylab("total debt") +
  ggtitle("Total Debt") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))

# debt/student over time
p2 = ggplot(df, aes(year, debt_per_student, group = school_no, col = school_no)) +
  geom_point() +
  geom_line() +
  ylab("debt per student") +
  ggtitle("Debt per Student") +
  scale_y_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        plot.title = element_text(size = 12))

plot_grid(p1, p2)
```

The total debt per schools fluctuates a lot over time. When we normalize by average daily membership---or total student enrollment---we see that debt per student stays relatively constant. However, there are some schools where debt per student spikes in the 2017-18 academic year. What happened here? Was there a policy change?

I also looked at the mean and variance of debt per student by school. This gives us a sense of \textit{how} much debt per student fluctuates within in each school. Schools with a high variance should be looked into---perhaps fluctuations in debt are tied to CEP status? Or someone bailing a school out of debt? Similarly, schools with a generally high mean debt per student are of interest. Why are these schools struggling more than others?


```{r mean_var, fig.width=8, fig.height=4}
# calculate mean and variance of debt/student by school
sums = df %>%
  group_by(school_no) %>%
  summarise(mean_debt = mean(debt_per_student, na.rm = TRUE),
            var = var(debt_per_student, na.rm = TRUE)) %>%
  arrange(-var)

# histogram of mean debt per student
h1 = ggplot(sums, aes(mean_debt)) +
  geom_histogram() +
 # ggtitle("Mean Debt-per-Student, 2010-11 through 2017-18") +
  xlab("mean") +
  scale_x_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        plot.margin = margin(2, 0, 2, 2, "cm"))

# histogram of variance of mean debt per student
h2 = ggplot(sums, aes(var)) +
  geom_histogram() +
 # ggtitle("Within-School Variance of Debt-per-Student, 2010-11 through 2017-18") + 
  xlab("variance") +
  scale_x_continuous(labels = scales::dollar) +
  theme(legend.position = "none",
        plot.margin = margin(1, 0, 2, 2, "cm"))

plot_grid(h1, h2,
          labels = c("Mean Debt-per-Student","Within-School Variance"),
          label_size = 12)
```

There are definitely some outliers in our data when it comes to variance of within-school debt per student and mean debt-per-student. Let's pull any schools that fall greater than 2 standard deviations from the mean (proper outliers) and see what their debt-per-student looks like longitudinally.

```{r}
# grab school codes for high variance
high.var = sums$school_no[sums$var > 100]

# debt/student over time
ggplot(df[df$school_no %in% high.var,], aes(year, debt_per_student, group = school_no, col = school_no)) +
  geom_point() +
  geom_line() +
  xlab("unpaid lunch debt") +
  ggtitle("Unpaid Lunch Debt per Student by School") +
  scale_y_continuous(labels = scales::dollar) +
  scale_fill_manual(labels = names)
```

This is pretty weird. Could it be something to do with the dataset?

Let's also look at schools with outlier values for mean debt.

```{r}
# calculate z-scores for mean debt per student
sums$z_debt = scale(sums$mean_debt, center = TRUE, scale = TRUE)

high.debt = sums$school_no[sums$z_debt >= 2]

# z-score greater than 2 is an outlier, plot those schools over time
ggplot(df[df$school_no %in% high.debt,], aes(year, debt_per_student, group = school_no, col = school_no)) +
  geom_point() +
  geom_line() +
  xlab("unpaid lunch debt") +
  ggtitle("Unpaid Lunch Debt per Student by School") +
  scale_y_continuous(labels = scales::dollar) +
  scale_fill_manual(labels = names)
```

- get 2010-11, 2011-12, 2012-13, 2013-14 and 2017-18 CEP data.
- get demographic data for 2016-17 and 2017-18 academic years
- double check datasets for conflicting ADM/counts

- think through when data is being pulled
- external factors (paying off debt, CEP status)
  - how to track people paying off debt/dates?
- think about fairly comparing schools
- LONG TERM: can a student graduate with debt? are they barred from anything? what are consequences besides food?